Concurrent Programming with CXL

### 2020-06-05 book/paper.tex
Improving paper

A good way to prove mutual exclusion is through induction on
the number of steps.
In such a proof, we need a so-called
\emph{inductive invariant} $I$, which is some predicate that holds
over the state.  An inductive invariant $I$, by definition, satisfies the following:
\begin{itemize}
\item (ME1) $I$ implies mutual exclusion.
\item (ME2) $I$ hold in the initial state.
\item (ME3) If $I$ holds over some state $s$, and some process in the
state takes a step, then $I$ also holds in the resulting state.
\end{itemize}

One candidate for such an invariant is mutual exclusion itself.
After all, it certainly satisfies ME1 and ME2.
And as CXL has already determined, mutual exclusion is an invariant:
it holds over every \emph{reachable state}.
Unfortunately, it is not an \emph{inductive} invariant and therefore does
not satisfy ME3.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, violating mutual exclusion.
So mutual exclusion is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

Let $C(\mathtt{self}) = \mathtt{flags}[1 - \mathtt{self}] \land
\mathtt{turn} = 1 - \mathtt{self}$, that is, the condition on the \texttt{while} loop.
For Peterson's Algorithm, an inductive invariant that works well is:
if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $C(i)$ does not hold or process $1-i$ is executing after setting
$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
Let's call it $I_p(i)$.

@@

To see that $I_p(i)$ implies mutual exclusion, suppose not.  Then
when both process~0 and process~1 are in the critical section, the
following must hold:
$(\lnot C(0) \lor \mathtt{process}(1)@gate) \land
 (\lnot C(1) \lor \mathtt{process}(0)@gate)$.
We know that both flags are set.
We also know that neither process~0 nor process~1 is at label \texttt{@gate}
(because they are both at label \texttt{@cs}),
so this simplifies to $(\lnot C(0)) \land (\lnot C(1))$.
So we conclude that $\mathtt{turn} = 0 \land \mathtt{turn} = 1$, a
logical impossibility.  Thus $I_p(i)$ satisfies ME1.

$I_p(i)$ satisfies ME2 trivially, because in the initial state no process
is in the critical section.

Finally we have to show that $I_p(i)$ satisfies ME3. Without loss of
generality, suppose $i=0$ (a benefit from the fact that the algorithm is
symmetric for both processes).  We have to show that if we are in a state
in which $I_p(0)$ holds, then any step will result in a state in which
$I_p(0)$ continues to hold.
Since $I_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
were to take a step, then in the next state process~0 would be no longer
at that label and $I_p(0)$ would hold trivially over the next state.
Therefore we only need to consider a step by process~1.

From $I_p(0)$ we know that one of the following cases must hold before
process~1 takes a step:
\begin{itemize}
\item \texttt{flags[1] = False};
\item \texttt{turn = 0};
\item process~1 is at label \texttt{@gate}.
\end{itemize}

Let us consider each of these cases.
In the first case, if process~1 takes a step, there are two possibilities:
either $flags[1]$ will still be \texttt{False} (in which case the first case
continues to hold), or $flags[1]$ will be \texttt{True}
and process~1 will be at label \texttt{@gate} (in which case the third case
will hold).
We know that process~1 never sets \texttt{turn} to 1, so
if the second case holds before the step, it will also hold after the step.
Finally, if process~1 is at label \texttt{@gate} before the step, then after
the step \texttt{turn} will equal 0, and therefore the second case will hold
after the step.

@@

A cool anecdote is the following.  When the author of CXL had to teach
Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
page.  The page claimed that the following predicate was invariant:
if process $i$ is in the critical section, then $\lnot C(i)$ (i.e.,
$I_p(i)$ without the disjunct that process $1-i$ is at label \texttt{gate}).
To demonstrate that this predicate is not invariant, you can remove the
disjunct from Figure~\ref{fig:petersonproof} and run it to get a
counterexample.

@@

A method \texttt{m} with argument \texttt{a} is invoked in its
most basic form as follows (assigning the result to \texttt{r}).
\begin{code}
\begin{verbatim}
r = m a;
\end{verbatim}
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
as \texttt{a}.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.
If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
the tuple consisting of values \texttt{a} and \texttt{b}.
You have already seen examples of this: the expression
\texttt{atLabel.cs} invokes method \texttt{atLabel} with
argument \texttt{.cs}.

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
to shared variables.
If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
(the parentheses are mandatory) is a pointer to \texttt{x}
(also known as the \emph{address} of \texttt{x}.
Conversely, if \texttt{p} is a pointer to a shared variable, then the
expression \texttt{\^{}p} is the value of the shared variable.

@@

Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple processes contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple processes.  Instead, processors provide
so-called \emph{interlock instructions}: special machine instructions
that can read memory and then write it in an indivisible step.

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language through its \texttt{atomic}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{verbatim}
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{verbatim}
\end{code}

Many CPUs have an atomic ``test-and-set'' operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here \texttt{s} points to a shared boolean variable and \texttt{p}
to a private boolean variable, belonging to some process.
The operation copies the value of the shared variable to the
private variable and then sets the shared variable to \texttt{True}.

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
It uses $N+1$ variables.
Variable \texttt{shared} is initialized to \texttt{False} while
\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
An important invariant, $I_1$, of the program is that at any time at most
one of these variables is \texttt{False}.
Another invariant, $I_2$, is that if process $i$ is in the critical section,
then \texttt{private[$i$] == False}.
Between the two, it is clear that only one process can be in the
critical section at the same time.

To see that invariant $I_1$ is maintained, note that
\texttt{\^{}p == True} upon entry of \texttt{tas}.
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
the invariant.
\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $I_1$ is also easy to verify for exiting the critical section.
Invariant $I_2$ is obvious as (i) process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False}, and (ii) no other process modifies
\texttt{private[$i$]}.

@@

CXL can check these invariants as well.  Figure~\ref{fig:tas} already
has the code to check $I_2$.  But how would one go about checking an
invariant like $I_1$.  Invariants must hold for every state.
For $I_2$ we only need an assertion at label \texttt{@cs} because the
premisse is that there is a process at that label.  However, we would
like to check $I_1$ in \emph{every state} (after the variables have
been initialized).

>>>>

A good way to prove mutual exclusion is through induction on
the number of steps.
In such a proof, we need a so-called \emph{inductive invariant} $I$ that
implies mutual exclusion.
An inductive invariant $I$ is a predicate over state that satisfies the following:
\begin{itemize}
\item $I$ hold in the initial state.
\item For any state in which $I$ holds and any process in the
state takes a step, then $I$ also holds in the resulting state.
\end{itemize}

One candidate for such an invariant is mutual exclusion itself.
After all, it certainly holds over the initial state.
And as CXL has already determined, mutual exclusion is an invariant:
it holds over every \emph{reachable state}.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, violating mutual exclusion.
So mutual exclusion is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

Let $C(i) = \mathtt{flags}[1 - i] \land
\mathtt{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
for process~$i$.
Let predicate $I_p(i)$ be the following.
if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $C(i)$ does not hold or process $1-i$ is executing after setting
$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
More formally, $I_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot C(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$I_p(0) \land I_p(1)$.

@@

To see that the inductive invariant implies mutual exclusion, suppose not.  Then
when both process~0 and process~1 are in the critical section, the
following must hold:
$(\lnot C(0) \lor \mathtt{process}(1)@gate) \land
 (\lnot C(1) \lor \mathtt{process}(0)@gate)$.
We know that both flags are set.
We also know that neither process~0 nor process~1 is at label \texttt{@gate}
(because they are both at label \texttt{@cs}),
so this simplifies to $(\lnot C(0)) \land (\lnot C(1))$.
So we conclude that $\mathtt{turn} = 0 \land \mathtt{turn} = 1$, a
logical impossibility.  Thus $I_p(i)$ implies mutual exclusion.

To see that $I_p(0) \land I_p(1)$ is, in fact, an inductive invariant, first note that
it certainly holds in the initial state, because in the initial state no process
is in the critical section.
Without loss of generality, suppose $i=0$ (a benefit from the fact that the algorithm is
symmetric for both processes).  We still have to show that if we are in a state
in which $I_p(0)$ holds, then any step will result in a state in which
$I_p(0)$ still holds.
If $I_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
were to take a step, then in the next state process~0 would be no longer
at that label and $I_p(0)$ would hold trivially over the next state.
Therefore we only need to consider a step by process~1.

From $I_p(0)$ we know that one of the following three cases must hold before
process~1 takes a step:
\begin{enumerate}
\item \texttt{flags[1] = False};
\item \texttt{turn = 0};
\item process~1 is at label \texttt{@gate}.
\end{enumerate}

Let us consider each of these cases.
In the first case, if process~1 takes a step, there are two possibilities:
either $flags[1]$ will still be \texttt{False} (in which case the first case
continues to hold), or $flags[1]$ will be \texttt{True}
and process~1 will be at label \texttt{@gate} (in which case the third case
will hold).
We know that process~1 never sets \texttt{turn} to 1, so
if the second case holds before the step, it will also hold after the step.
Finally, if process~1 is at label \texttt{@gate} before the step, then after
the step \texttt{turn} will equal 0, and therefore the second case will hold
after the step.  qed.

@@

A cool anecdote is the following.  When the author of CXL had to teach
Peterson's Algorithm, he refreshed his memory by looking at the Wikipedia
page.  The page claimed that the following predicate is invariant:
if process $i$ is in the critical section, then $\lnot C(i)$ (i.e.,
$I_p(i)$ without the disjunct that process $1-i$ is at label \texttt{gate}).
To demonstrate that this predicate is not invariant, you can remove the
disjunct from Figure~\ref{fig:petersonproof} and run it to get a
counterexample.

@@

A method \texttt{m} with argument \texttt{a} is invoked in its
most basic form as follows (assigning the result to \texttt{r}).
\begin{code}
\begin{verbatim}
r = m a;
\end{verbatim}
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
as \texttt{a}.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.
If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
the tuple consisting of values \texttt{a} and \texttt{b}.

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
to shared variables.
If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
(the parentheses are mandatory) is a pointer to \texttt{x}
(also known as the \emph{address} of \texttt{x}).
Conversely, if \texttt{p} is a pointer to a shared variable, then the
expression \texttt{\^{}p} is the value of the shared variable.

@@

Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple processes contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple processes.  Instead, multi-core processors provide
so-called \emph{interlock instructions}: special machine instructions
that can read memory and then write it in an indivisible step.

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{verbatim}
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{verbatim}
\end{code}

Many CPUs have an atomic ``test-and-set'' operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here \texttt{s} points to a shared boolean variable and \texttt{p}
to a private boolean variable, belonging to some process.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to \texttt{True} (``set'').

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
It uses $N+1$ variables.
Variable \texttt{shared} is initialized to \texttt{False} while
\texttt{private[$i$]} for each process $i$ is initialized to \texttt{True}.
An important invariant, $I_1$, of the program is that at any time at most
one of these variables is \texttt{False}.
Another invariant, $I_2(i)$, is that if process $i$ is in the critical section,
then \texttt{private[$i$] == False}.
Between the two (i.e., $I_1 \land \forall i: I_2(i)$),
it is clear that only one process can be in the
critical section at the same time.

To see that invariant $I_1$ is maintained, note that
\texttt{\^{}p == True} upon entry of \texttt{tas}.
So there are two cases:
\begin{enumerate}
\item \texttt{\^{}s} is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{\^{}p == False} and \texttt{\^{}s == True}, maintaining
the invariant.
\item \texttt{\^{}s} is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $I_1$ is also easy to verify for exiting the critical section.
Invariant $I_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \texttt{private[$i$] == False}, and (ii) no other process modifies
\texttt{private[$i$]}.

@@

CXL can check these invariants as well.  Figure~\ref{fig:tas} already
has the code to check $I_2(i)$.  But how would one go about checking an
invariant like $I_1$.  Invariants must hold for every state.
For $I_2$ we only need an assertion at label \texttt{@cs} because the
premisse is that there is a process at that label.  However, we would
like to check $I_1$ in \emph{every state} (after the variables have
been initialized).
